New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-16 jso.qmd
minor typos

To sample bases for calculating smoothness, four arguments are required: 1) the index function, the dataset, the number of random bases to sample, and the best basis to compute projection distance. The output of `sample_bases()` is a data frame containing columns/ list-columns of sampled basis matrix, calculated projection distance and index value. Parallellasation is available through the `parallel = TRUE` argument to speed up the computation.

@@

The `calc_squintability()` function computes squintability using two methods: 1) parametrically, by fitting a scaled sigmoid function through non-linear least square (`method = "nls"`), and 2) non-parametrically, using kernel smoothing (`method = "ks"`). A `bin_width` argument is required to average the index values over the projection distance before the fit. The output provides the estimated parameters for the scaled sigmoid function ($\theta_1$ to $\theta_4$) and the squintability metric for the parametric case. For the non-parametric case, it shows the maximum gradient attained(`max_d`), the corresponding projection distance (`max_dist`), and the squintability metric as their products.

>>>>

To sample bases for calculating smoothness, four arguments are required: the index function, the dataset, the number of random bases to sample, and the best basis to compute projection distance. The output of `sample_bases()` is a data frame containing columns/ list-columns of sampled basis matrix, calculated projection distance and index value. Parallellasation is available through the `parallel = TRUE` argument to speed up the computation.

@@

The `calc_squintability()` function computes squintability using two methods: 1) parametrically, by fitting a scaled sigmoid function through non-linear least square (`method = "nls"`), and 2) non-parametrically, using kernel smoothing (`method = "ks"`). A `bin_width` argument is required to average the index values over the projection distance before the fit. The output provides the estimated parameters for the scaled sigmoid function ($\theta_1$ to $\theta_4$) and the squintability metric for the parametric case. For the non-parametric case, it shows the maximum gradient attained (`max_d`), the corresponding projection distance (`max_dist`), and the squintability metric as their products.

### 2024-09-16 jso.tex
minor typos

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, boxrule=0pt, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, breakable, coltitle=black, opacitybacktitle=0.6, arc=.35mm, toprule=.15mm, left=2mm, bottomtitle=1mm, leftrule=.75mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}]

@@

To sample bases for calculating smoothness, four arguments are required:
1) the index function, the dataset, the number of random bases to
sample, and the best basis to compute projection distance. The output of
\texttt{sample\_bases()} is a data frame containing columns/
list-columns of sampled basis matrix, calculated projection distance and
index value. Parallellasation is available through the
\texttt{parallel\ =\ TRUE} argument to speed up the computation.

@@

The \texttt{calc\_squintability()} function computes squintability using
two methods: 1) parametrically, by fitting a scaled sigmoid function
through non-linear least square (\texttt{method\ =\ "nls"}), and 2)
non-parametrically, using kernel smoothing (\texttt{method\ =\ "ks"}). A
\texttt{bin\_width} argument is required to average the index values
over the projection distance before the fit. The output provides the
estimated parameters for the scaled sigmoid function (\(\theta_1\) to
\(\theta_4\)) and the squintability metric for the parametric case. For
the non-parametric case, it shows the maximum gradient
attained(\texttt{max\_d}), the corresponding projection distance
(\texttt{max\_dist}), and the squintability metric as their products.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacityback=0, opacitybacktitle=0.6, colback=white, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, arc=.35mm, toprule=.15mm, rightrule=.15mm, left=2mm, bottomtitle=1mm]

@@

To sample bases for calculating smoothness, four arguments are required:
the index function, the dataset, the number of random bases to sample,
and the best basis to compute projection distance. The output of
\texttt{sample\_bases()} is a data frame containing columns/
list-columns of sampled basis matrix, calculated projection distance and
index value. Parallellasation is available through the
\texttt{parallel\ =\ TRUE} argument to speed up the computation.

@@

The \texttt{calc\_squintability()} function computes squintability using
two methods: 1) parametrically, by fitting a scaled sigmoid function
through non-linear least square (\texttt{method\ =\ "nls"}), and 2)
non-parametrically, using kernel smoothing (\texttt{method\ =\ "ks"}). A
\texttt{bin\_width} argument is required to average the index values
over the projection distance before the fit. The output provides the
estimated parameters for the scaled sigmoid function (\(\theta_1\) to
\(\theta_4\)) and the squintability metric for the parametric case. For
the non-parametric case, it shows the maximum gradient attained
(\texttt{max\_d}), the corresponding projection distance
(\texttt{max\_dist}), and the squintability metric as their products.
